---
title: "STAT545a_Class04"
author: "Adrian jones"
date: "September 17, 2015"
output: html_document
---

##We are using the gapminder library way to work on stuff

First we'll initialise the world and load in gapminder

```{r}
rm(list = ls())

library(gapminder)
library(ggplot2)

str(gapminder)
````

This should show a data.frame object.  Like we know, they are supper useful things b/c we can put anything we want in them.

str is a useful function that shows the structure of an object, including telling you the dimension of things, what type of variable (factor, numeric, string etc) is in each column etc.

```{r}
head(gapminder)

tail(gapminder)

gapminder[sample(nrow(gapminder), 10),]
```

Looking at the head and tail and a random sample of the data set reasures us that the data loaded happily.  Particularly the random sample is clever.

sample size 10, x is number of rows in gapminder, then it is a normal data.frame assignment

##Lets plot some Graphs

Because graphs are always good

```{r}
plot(lifeExp ~ year, data = gapminder)

plot(lifeExp ~ gdpPercap, data = gapminder)
```

Jennifer did something interesting by using the formual call instead of the usual x,y call of plot.   But it gave the same thing.  Is it better?  I dunno.

So with the gdp plot, probably a log scale would show things to better effect.

```{r}
plot(lifeExp ~ log(gdpPercap), gapminder)
```

##Now we should probably have a look at individual variables

###lifeExp

```{r}
head(gapminder$lifeExp)
tail(gapminder$lifeExp)
summary(gapminder$lifeExp)
hist(gapminder$lifeExp)

```

We grind through variables to see what they are like, histogram, density plot ext

###What about year?

```{r}
summary(gapminder$year)
class(gapminder$year)
#But it isn't a great nummeric

table(gapminder$year)
```

###continent
Its a factor, we can check that with the 

> class

Function

```{r}
class(gapminder$continent)
levels(gapminder$continent)
table(gapminder$continent)
```

#### A side comment on Factors
Intesrestingly when something is a factor it is alwayus stored as numerics and then tied to a value for that factor.  Ie in this situation continents are 1:5

R did this automatically when it read the data as a table.
That behaviour can be supressed with the argument

> strings.as.factor=FALSE

in the various read.XX functions.

Tools in the Hadelyverse tend to avoid this problem as default


####So back to the continents
We still need a plot

```{r}
barplot(table(gapminder$continent))
```
## Halajuya it is ggplot2 time

```{r}
library(ggplot2)
p <- ggplot(subset(gapminder, continent != "Oceania"),
            aes(x = gdpPercap, y = lifeExp)) # just initializes
p <- p + scale_x_log10() # log the x axis the right way
p + geom_point() # scatterplot
p + geom_point(aes(color = continent)) # map continent to color
p + geom_point(alpha = (1/3), size = 3) + geom_smooth(lwd = 3, se = FALSE)
## geom_smooth: method="auto" and size of largest group is >=1000, so using gam with formula: y ~ s(x, bs = "cs"). Use 'method = x' to change the smoothing method.
p + geom_point(alpha = (1/3), size = 3) + facet_wrap(~ continent)
```
> How does this stuff work?  I have no clue and should figure it out

##OTher options including subsets of stuff
Lots of things have a argument

> subset

Which lets you subset based upon things other than matrix indicies, which is a really good thing when you don't want to have to comment the heck out of your stuff later on

```{r}
subset(gapminder, subset = country == "Uruguay")
```

